---
output: github_document
bibliography: ref.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(digits = 3)
```

# mdgc

This package contains a marginal likelihood approach to estimating the 
model discussed by @hoff07 and @zhao19. We have modified the Fortran
code by @Genz02 to supply an approximation gradient for the log 
marginal likelihood and to use an approximation of the marginal likelihood 
similar to the CDF approximation in @Genz02.

However, the package can be useful for a lot of other models. For instance, 
the methods are directly applicable to other Gaussian copula models and some 
mixed effect models. All methods are implemented in C++, support computation 
in parallel, and should easily be able to be ported to other languages. 

## Example
Below, we provide an example similar to @zhao19 [Section 7.1]. The authors 
use a data set with a random correlation matrix, 5 continuous variables, 
5 binary variables, and 5 ordinal variables with 5 levels. There is a total 
of 2000 observations and 30% of the variables are missing completely at 
random. 

To summarize @zhao19 results, they show that their approximate EM algorithm 
converges in what seems to be 20-25 seconds while it takes more than 150
seconds for the MCMC algorithm used by @hoff07. In essence, these figures
should be kept in mind when looking at the results below. Importantly, 
@zhao19 use an approximation in the E-step of the EM algorithm which is 
fast but might be crude is some settings. Using a potentially arbitrarily 
precise approximation of the marginal likelihood is useful if this can 
be done quickly enough. 

To summarize, we do the following below: 

 1. simulate the data set we will use. 
 2. show how to use the C++ functions and that these provide an 
    approximation of the log marginal likelihood and its gradient. Moreover, 
    we show that the methods scales well in the number of threads.
 3. define functions to perform maximum likelihood estimation. 
 4. estimate the parameters using a simple gradient descent algorithm and 
    stochastic gradient descent using ADAM. 
 5. show how to improve 4. by using better starting values which are quick 
    to compute. As of this writing, this reduces the estimation time to 
    less than 5 seconds using four threads and less than 13.5 seconds using 
    one thread.
    
Presumably/likely, computing the marginals of each variable should be 
extremely fast, as should 
the imputation once the model parameters are estimated. Given that this is 
true, then the main concern should be the time it takes to estimate the 
model parameters. As we show, this can be done quickly using our code.

```{r sim_dat, cache = 1, fig.height = 3, fig.width = 7}
library(mdgc)

# simulates a data set and puts it into a format to pass to C++.
#
# Args: 
#   n: number of observations. 
#   p: number of variables. 
#   n_lvls: number of levels for the ordinal variables. 
# 
# Returns: 
#   Simluated masked data set in the format needed to pass to C++.
sim_dat <- function(n, p = 4, n_lvls = 5L){
  # get the covariance matrix
  Sb <- diag(p)
  Sb[lower.tri(Sb)] <- Sb[upper.tri(Sb)] <- .5
  Sb <- Sb / p / 5
  Sig <- cov2cor(drop(rWishart(1L, 5L * p, Sb)))
  
  # draw the observations
  Z <- truth <- crossprod(chol(Sig), matrix(rnorm(n * p), p))
  
  # determine the type
  type <- rep(1:3, each = floor((p + 3 - 1) / 3))[1:p]
  is_con <- type == 1L
  is_bin <- type == 2L
  is_ord <- type == 3L
  
  # mask data 
  is_mask <- matrix(runif(n * p) < .3, p)
  Z[is_mask] <- NA_real_
  
  # set code arugment. Codes are:
  #  0: latent Z is observed (upper is the observed point).
  #  1: latent Z can be anything.
  #  2: latent Z is in an interval given by lower and upper.
  code <- matrix(0L, p, n)
  code[is_ord | is_bin, ] <- 2L
  code[is_mask] <- 1L
  
  # create upper and lower bounds
  upper <- lower <- matrix(NA_real_, p, n)
  lower[is_con, ] <- upper[is_con, ] <- truth[is_con, ]
  
  bs <- c(-Inf, 0, Inf)
  lower[is_bin, ] <- head(bs, -1)[cut(truth[is_bin, ], breaks = bs)]
  upper[is_bin, ] <- bs[-1]      [cut(truth[is_bin, ], breaks = bs)]
  
  bs <- qnorm(seq(0, 1, length.out = n_lvls + 1L))
  lower[is_ord, ] <- head(bs, -1)[cut(truth[is_ord, ], breaks = bs)]
  upper[is_ord, ] <- bs[-1]      [cut(truth[is_ord, ], breaks = bs)]
  
  lower[is_mask] <- upper[is_mask] <- NA_real_
  
  list(lower = lower, upper = upper, code = code, Sigma = Sig, 
       truth = truth)
}

# simulate and show the data. Data is stored in 
# [# variables] x [# observations] format
set.seed(1)
p <- 15L
dat <- sim_dat(2000L, p = p)
dat$lower[, 1:5] # lower bounds
dat$upper[, 1:5] # upper bounds
dat$code [, 1:5] # code for type of data
dat$truth[, 1:5] # true values

#####
# get pointers to objects in C++
system.time(
  ptr <- mdgc:::get_log_lm_terms(lower = dat$lower, upper = dat$upper, 
                                 code = dat$code))

# approximates the log marginal likelihood function.
#
# Args: 
#   vcov: correlation matrix.
#   releps: relative error for each term. 
#   n_threads: number of threads. 
#   comp_derivs: logical for whether to approximate the gradient. 
#   indices: integer vector with which terms to include. 
#   do_reorder: logical for whether to use heuristic variable reordering.
log_ml <- function(vcov, releps = 1e-2, n_threads = 4L, comp_derivs = FALSE, 
                   indices = 0:(NCOL(dat$lower) - 1L), do_reorder = TRUE)
  mdgc:::eval_log_lm_terms(
    ptr = ptr, vcov = vcov, indices = indices, 
    maxpts = 100000L, abseps = -1, releps = releps, n_threads = n_threads, 
    comp_derivs = comp_derivs, do_reorder = do_reorder)

# print the approximate log marginal likelihood at the true parameters
set.seed(1)
print(log_ml(dat$Sigma, n_threads = 1L), digits = 7)

# check standard error
sd(replicate(20, log_ml(dat$Sigma, n_threads = 4L)))

# without reordering
print(log_ml(dat$Sigma, n_threads = 4L, do_reorder = FALSE), digits = 7)

# check standard error
sd(replicate(20, log_ml(dat$Sigma, n_threads = 4L, do_reorder = FALSE)))

# check computation time
library(microbenchmark)
microbenchmark(
  `1 thread                 ` = 
    log_ml(dat$Sigma                    , n_threads = 1L), 
  `1 thread  (w/o rordering)` = 
    log_ml(dat$Sigma, do_reorder = FALSE, n_threads = 1L), 
  `2 threads                ` = 
    log_ml(dat$Sigma                    , n_threads = 2L),
  `2 threads (w/o rordering)` = 
    log_ml(dat$Sigma, do_reorder = FALSE, n_threads = 2L),
  `4 threads                ` = 
    log_ml(dat$Sigma                    , n_threads = 4L), 
  `4 threads (w/o rordering)` = 
    log_ml(dat$Sigma, do_reorder = FALSE, n_threads = 4L), 
  times = 5)

#####
# we can also get an approximation of the gradient
t1 <- log_ml(dat$Sigma, comp_derivs = TRUE, n_threads = 1L)
t2 <- log_ml(dat$Sigma, comp_derivs = TRUE, n_threads = 4L)
all.equal(t1, t2, tolerance = 1e-2)
  
microbenchmark(
  `1 thread                 ` = 
    log_ml(dat$Sigma, comp_derivs = TRUE                    , n_threads = 1L), 
  `1 thread  (w/o rordering)` = 
    log_ml(dat$Sigma, comp_derivs = TRUE, do_reorder = FALSE, n_threads = 1L), 
  `2 threads                ` = 
    log_ml(dat$Sigma, comp_derivs = TRUE                    , n_threads = 2L),
  `2 threads (w/o rordering)` = 
    log_ml(dat$Sigma, comp_derivs = TRUE, do_reorder = FALSE, n_threads = 2L),
  `4 threads                ` = 
    log_ml(dat$Sigma, comp_derivs = TRUE                    , n_threads = 4L), 
  `4 threads (w/o rordering)` = 
    log_ml(dat$Sigma, comp_derivs = TRUE, do_reorder = FALSE, n_threads = 4L), 
  times = 5)

#####
# the main code in the packages provides an approximation to the CDF similar 
# to the one in the mvtnorm package. We provide an example below to 
# illustrate this. Feel free to skip this part of the README.
library(mvtnorm)
set.seed(1)
p_ex <- 5L
S_ex <- diag(p_ex)
S_ex[lower.tri(S_ex)] <- S_ex[upper.tri(S_ex)] <- .25
m_ex <- seq(-2, 2, length.out = p_ex)
lower_ex <- m_ex + drop(rnorm(p_ex) %*% chol(S_ex)) - 1
upper_ex <- lower_ex + 1

use_mvtnorm <- function()
  pmvnorm(
    lower = lower_ex, upper = upper_ex, sigma = S_ex, mean = m_ex, 
    algorithm = GenzBretz(maxpts = 10000L, abseps = -1, releps = 1e-6))
use_this_pkg <- function(derivs = FALSE)
  mdgc:::pmvnorm(lower = lower_ex, upper = upper_ex, mu = m_ex, 
                 Sigma = S_ex, maxvls = 10000L, abseps = -1, releps = 1e-6, 
                 derivs = derivs)
use_mvtnorm()
use_this_pkg()
all.equal(c(use_mvtnorm()), c(use_this_pkg()), tolerance = 1e-5)
microbenchmark(mvtnorm = use_mvtnorm(), mdgc = use_this_pkg(), 
               times = 25)

sd(replicate(25, use_mvtnorm()))
sd(replicate(25, use_this_pkg()))

# the latter function can also provide gradients with respect to the mean 
# and covariance matrix
library(numDeriv)
gr_hat <- jacobian(function(a){
  m <- a[1:p_ex]
  S <- matrix(nr = p_ex, nc = p_ex)
  S[upper.tri(S, TRUE)] <- a[-(1:p_ex)]
  S[lower.tri(S)] <- t(S)[lower.tri(S)]
  
  set.seed(1L)
  res <- pmvnorm(
    lower = lower_ex, upper = upper_ex, sigma = S, mean = m, 
    algorithm = GenzBretz(maxpts = 10000L, abseps = -1, releps = 1e-6))
  c(res)
}, c(m_ex, S_ex[upper.tri(S_ex, TRUE)]))
gr <- use_this_pkg(TRUE)

# the off diagonal elements of the covariance matrix is not scaled by 2
gr_hat / gr[-1]

# creates a matrix from a log-Cholesky decomposition.
# 
# Args:
#   par: p (p + 1) / 2 elements in the log-Cholesky decomposition.
get_lchol_inv <- function(par){
  # use log-cholesky parametrization
  p <- (sqrt(8 * length(par) + 1) - 1) / 2
  L <- matrix(0, p, p)
  L[lower.tri(L, TRUE)] <- par
  diag(L) <- exp(diag(L))
  tcrossprod(L)
}

# creates the log-Cholesky decomposition. 
# 
# Args: 
#   par: positive definite matrix to decompose
get_lchol <- function(par){
  lSig <- t(chol(par))
  diag(lSig) <- log(diag(lSig))
  lSig[lower.tri(lSig, TRUE)]
}

# indices used to apply a matrix product with a get_commutation matrix
com_vec <- mdgc:::get_commutation_vec(p, p, FALSE)

# computes the approximate log marginal likelihood. 
#
# Args:
#   par: log-Cholesky decomposition.
#   seed: seed to use.  
#   comp_derivs: logical for whether to approximate the gradient. 
#   n_threads: number of threads. 
#   releps: relative error for each term.
#   indices: integer vector with which terms to include. 
par_fn <- function(par, seed = NULL, comp_derivs = FALSE, 
                   n_threads = 1L, releps = 1e-2, 
                   indices = 0:(NCOL(dat$lower) - 1L)){
  if(!is.null(seed))
    set.seed(seed)
  Arg <- get_lchol_inv(par)
  
  res <- log_ml(Arg, comp_derivs = comp_derivs, indices = indices,
                n_threads = n_threads, releps = releps)
  log_ml <- c(res)
  if(comp_derivs){
    gr <- attr(res, "grad")
    tmp <- matrix(0, p, p)
    tmp[lower.tri(tmp, TRUE)] <- par
    diag(tmp) <- exp(diag(tmp))
    gr <- gr[com_vec] + c(gr)
    gr <- mdgc:::x_dot_X_kron_I(x = gr, X = tmp, l = p)
    gr <- gr[, lower.tri(tmp, TRUE)]
    idx_diag <- c(1L, 1L + cumsum(NCOL(tmp):2)) 
    gr[idx_diag] <- gr[idx_diag] * diag(tmp)
      
    attr(log_ml, "grad") <- gr
    
  }
  
  log_ml
}

# check that the function gives the correct log marginal likelihood
# approximation and gradient approximation.
lSig <- get_lchol(dat$Sigma)
r1 <- par_fn(lSig, comp_derivs = TRUE, n_threads = 4L, releps = 1e-3, 
             indices = 1:100)
r2 <- numDeriv::jacobian(par_fn, lSig, seed = 1L, n_threads = 6L, 
                         releps = 1e-3, indices = 1:100)
all.equal(attr(r1, "grad"), drop(r2), tolerance = 1e-2)

#####
# performs gradient descent. 
# 
# Args: 
#   val: starting value. 
#   step_start: starting value for the step length. 
#   n_threads: number of threads to use. 
#   maxit: maximum number of iteration. 
#   eps: convergence threshold to use. 
#   seed: seed to use.
naiv_gradient_descent <- function(val, step_start, n_threads = 4L, 
                                  maxit = 10L, eps = 1e-3, seed = 1L){
  fun_vals <- step_sizes <- rep(NA_real_, maxit)
  
  for(i in 1:maxit){
    gr <- par_fn(val, comp_derivs = TRUE, n_threads = n_threads, 
                 seed = seed)
    fun_vals[i] <- prev_fun <- c(gr)
    dir <- attr(gr, "grad")
    step <- step_start
    if(sqrt(sum(dir^2)) < eps)
      break
    
    max_j <- 11L
    for(j in 1:max_j){
      if(j == max_j)
        warning("Failed to find a decrease")
      new_val <- val + step * dir
      new_val <- get_lchol(cov2cor(get_lchol_inv(new_val)))
      new_fun <- par_fn(new_val, comp_derivs = FALSE, n_threads = n_threads, 
                        seed = seed)
      if(new_fun > prev_fun){
        val <- new_val
        break
      }
      step <- step / 2
    }
    
    step_sizes[i] <- step
  }
  
  list(result = get_lchol_inv(val), logml = prev_fun, 
       nit = i, step_sizes = step_sizes, fun_vals = fun_vals)
}

# estimate model parameters
start_val <- numeric(p * (p + 1) / 2)
system.time(res <- naiv_gradient_descent(val = start_val, step_start = .001, 
                                         maxit = 20L, eps = 1e-2))

# compare estimates with truth
norm(res$result - dat$Sigma)
res$result
dat$Sigma

# or plot both of them and compare
do_plot <- function(x, main){
  sc <- colorRampPalette(c("Red", "White", "Blue"))(50)
  image(x, main = main, col = sc, zlim = c(-1, 1), xaxt = "n", yaxt = "n", 
        bty = "n")
}
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
do_plot(res$result, "Estimates")
do_plot(dat$Sigma, "Truth")
do_plot(res$result - dat$Sigma, "Difference")

res$fun_vals # log marginal likelihood estimates at each iteration

#####
# performs stochastic gradient descent instead (using ADAM).
# 
# Args: 
#   val: starting value. 
#   batch_size: number of observations in each batch. 
#   n_threads: number of threads to use. 
#   maxit: maximum number of iteration. 
#   seed: seed to use.
#   epsilon, alpha, beta_1, beta_2: ADAM parameters.
adam <- function(val, batch_size, n_threads = 4L, maxit = 10L, 
                 seed = 1L, epsilon = 1e-8, alpha = .001, beta_1 = .9, 
                 beta_2 = .999){
  indices <- sample(0:(NCOL(dat$lower) - 1L), replace = FALSE)
  blocks <- tapply(indices, (seq_along(indices) - 1L) %/% batch_size, 
                   identity, simplify = FALSE)
  
  n_blocks <- length(blocks)
  n_par <- length(val)
  m <- v <- numeric(n_par)
  fun_vals <- numeric(maxit)
  estimates <- matrix(NA_real_, n_par, maxit)
  i <- -1L
  
  for(k in 1:maxit){
    for(ii in 1:n_blocks){
      i <- i + 1L
      idx_b <- (i %% n_blocks) + 1L
      m_old <- m
      v_old <- v
      res <- par_fn(val, comp_derivs = TRUE, n_threads = n_threads, 
                    seed = seed, indices = blocks[[idx_b]])
      fun_vals[(i %/% n_blocks) + 1L] <- 
        fun_vals[(i %/% n_blocks) + 1L] + c(res)
      
      gr <- attr(res, "grad")
      
      m <- beta_1 * m_old + (1 - beta_1) * gr
      v <- beta_2 * v_old + (1 - beta_2) * gr^2
      
      m_hat <- m / (1 - beta_1^(i + 1))
      v_hat <- v / (1 - beta_2^(i + 1))
      
      val <- val + alpha * m_hat / (sqrt(v_hat) + epsilon)
      val <- get_lchol(cov2cor(get_lchol_inv(val)))
    }
    
    estimates[, k] <- val
  }
  
  list(result = get_lchol_inv(val), fun_vals = fun_vals, 
       estimates = estimates)
}

# estimate the model parameters
set.seed(1)
system.time(res_adam  <- adam(
  val = start_val, alpha = 1e-2, maxit = 10L, batch_size = 100L))

# compare estimates with the truth
norm(res_adam$result - dat$Sigma)
res_adam$result
dat$Sigma

# use plot instead
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
do_plot(res_adam$result, "Estimates (ADAM)")
do_plot(dat$Sigma, "Truth")
do_plot(res_adam$result - dat$Sigma, "Difference")

# look at the maximum log marginal likelihood both at the end and after 
# each iteration
log_ml(res_adam$result)
funvals_adam_org <- 
  apply(res_adam$estimates, 2L, function(x) log_ml(get_lchol_inv(x)))
funvals_adam_org
res_adam$fun_vals # likely lower bounds on the log-marginal likelihood
res_adam_org <- res_adam

#####
# performs stochastic gradient descent instead (using SVRG).
# 
# Args: 
#   val: starting value. 
#   batch_size: number of observations in each batch. 
#   n_threads: number of threads to use. 
#   maxit: maximum number of iteration. 
#   seed: seed to use.
#   lr: learning rate. 
svrg <- function(val, batch_size, n_threads = 4L, maxit = 10L, 
                 seed = 1L, lr){
  all_indices <- 0:(NCOL(dat$lower) - 1L)
  indices <- sample(all_indices, replace = FALSE)
  blocks <- tapply(indices, (seq_along(indices) - 1L) %/% batch_size, 
                   identity, simplify = FALSE)
  
  n_blocks <- length(blocks)
  n_par <- length(val)
  estimates <- matrix(NA_real_, n_par, maxit + 1L)
  fun_vals <- numeric(maxit + 1L)
  estimates[, 1L] <- val
  
  for(k in 1:maxit + 1L){
    old_val <- estimates[, k - 1L]
    old_grs <- sapply(1:n_blocks - 1L, function(ii){
      idx_b <- (ii %% n_blocks) + 1L
      res_old <- par_fn(old_val, comp_derivs = TRUE, n_threads = n_threads, 
                        seed = seed, indices = blocks[[idx_b]])
      c(res_old, attr(res_old, "grad"))
    })
    
    fun_vals[k - 1L] <- sum(old_grs[1, ])
    old_grs <- old_grs[-1L, , drop = FALSE ]
    old_gr <- rowSums(old_grs) / n_blocks
    
    for(ii in 1:n_blocks - 1L){
      idx_b <- (ii %% n_blocks) + 1L
      res <- par_fn(val, comp_derivs = TRUE, n_threads = n_threads, 
                    seed = seed, indices = blocks[[idx_b]])
      fun_vals[k] <- fun_vals[k] + c(res)
      dir <- attr(res, "grad") - old_grs[, ii + 1L] + old_gr
      
      val <- val + lr * dir
      val <- get_lchol(cov2cor(get_lchol_inv(val)))
    }
    
    estimates[, k] <- val
  }
  
  list(result = get_lchol_inv(val), fun_vals = fun_vals[-1L], 
       estimates = estimates[, -1L, drop = FALSE])
}

# estimate the model parameters
set.seed(1)
system.time(res_svrg  <- svrg(
  val = start_val, lr = 1e-3, maxit = 10L, batch_size = 100L))

# compare estimates with the truth
norm(res_svrg$result - dat$Sigma)
res_svrg$result
dat$Sigma

# use plot instead
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
do_plot(res_svrg$result, "Estimates (SVRG)")
do_plot(dat$Sigma, "Truth")
do_plot(res_svrg$result - dat$Sigma, "Difference")

# look at the maximum log marginal likelihood both at the end and after 
# each iteration
funvals_svrg_org <- res_svrg$fun_vals
funvals_svrg_org[length(funvals_svrg_org)] <- log_ml(res_svrg$result)
funvals_svrg_org

#####
# we can use better starting values. E.g. something heuristic like: 
#   - transform back into the [0, 1] scale. 
#   - take the middle of the interval and map back. 
#   - compute the partial correlations. 
get_z_hat <- function(lower, upper, code){
  out <- mapply(function(l, u, co){
    if(co <= 1)
      return(u)
    
    a <- if(is.infinite(l)) 0 else pnorm(l)
    b <- if(is.infinite(u)) 1 else pnorm(u)
    qnorm((a + b) / 2)
  }, l = lower, u = upper, c = code)
  dim(out) <- dim(lower)
  out
}
tmp <- get_z_hat(dat$lower, dat$upper, dat$code)

# we also have a C++ function to do this which is faster
all.equal(tmp, mdgc:::get_z_hat(
  dat$lower, dat$upper, dat$code, n_threads = 4L))

# the latter is faster but both are relatively fast
microbenchmark(
  `R version  ` = get_z_hat(dat$lower, dat$upper, dat$code), 
  `C++ verison` = mdgc:::get_z_hat(
  dat$lower, dat$upper, dat$code, n_threads = 4L), times = 10)

# then we can compute an approximation of the covariance matrix as follows
system.time(chat <- cov2cor(cov(t(tmp), use = "pairwise.complete.obs")))

# the starting value is already quite close
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
norm(chat - dat$Sigma)
do_plot(chat, "Starting value")
do_plot(dat$Sigma, "Truth")
do_plot(chat - dat$Sigma, "Difference")

# run ADAM again 
start_val <- get_lchol(chat)
set.seed(1)
system.time(res_adam  <- adam(
  val = start_val, alpha = 1e-2, maxit = 5L, batch_size = 100L))

# for comparisons, we also run the code using one thread
set.seed(1)
system.time(res_adam_ser  <- adam(
  val = start_val, alpha = 1e-2, maxit = 5L, batch_size = 100L, 
  n_threads = 1L))

# we get (roughly) the same
norm(res_adam$result - res_adam_ser$result)

# plot estimate
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
norm(res_adam$result - dat$Sigma)
do_plot(res_adam$result, "Estimates (ADAM)")
do_plot(dat$Sigma, "Truth")
do_plot(res_adam$result - dat$Sigma, "Difference")

# check log marginal likelihood like before
log_ml(res_adam$result)
funvals_adam <- 
  apply(res_adam$estimates, 2L, function(x) log_ml(get_lchol_inv(x)))
funvals_adam
res_adam$fun_vals # likely lower bounds on the log-marginal likelihood

# do the same with SVRG
set.seed(1)
system.time(res_svrg  <- svrg(
  val = start_val, lr = 1e-3, maxit = 5L, batch_size = 100L))

# compare estimates with the truth
norm(res_svrg$result - dat$Sigma)
res_svrg$result
dat$Sigma

# use plot instead
par(mfcol = c(1, 3), mar  = c(1, 1, 4, 1))
do_plot(res_svrg$result, "Estimates (SVRG)")
do_plot(dat$Sigma, "Truth")
do_plot(res_svrg$result - dat$Sigma, "Difference")

# look at the maximum log marginal likelihood both at the end and after 
# each iteration
funvals_svrg <- res_svrg$fun_vals
funvals_svrg[length(funvals_svrg)] <- log_ml(res_svrg$result)
funvals_svrg

#####
# compare convergence of the different methods 
#  line: gradient descent. 
#  dashed: ADAM with poor starting values. 
#  dotted: ADAM with better starting values
#  blue: same as ADAM but using SVRG.
lls <- matrix(NA_real_, max(length(res$fun_vals), length(funvals_adam_org), 
                            length(funvals_adam), length(funvals_svrg_org), 
                            length(funvals_svrg)), 5L)
lls[seq_along(res$fun_vals)    , 1] <- res$fun_vals
lls[seq_along(funvals_adam_org), 2] <- funvals_adam_org
lls[seq_along(funvals_adam)    , 3] <- funvals_adam
lls[seq_along(funvals_svrg_org), 4] <- funvals_svrg_org
lls[seq_along(funvals_svrg)    , 5] <- funvals_svrg

par(mfcol = c(1, 1), mar = c(5, 5, 1, 1))
matplot(
  lls, lty = c(1:3, 2:3), col = c(rep("black", 3), rep("darkblue", 2)), 
  bty = "l", type = "l", xlab = "Iteration", 
  ylab = "Log marginal likelihood")

# skipping the first n steps
n_skip <- 5L
matplot(
  lls, lty = c(1:3, 2:3), col = c(rep("black", 3), rep("darkblue", 2)), 
  ylim = range(lls[-(1:n_skip), ], na.rm = TRUE), bty = "l", 
  type = "l", xlab = "Iteration", ylab = "Log marginal likelihood")
```

## References
